---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Laura Fontanills"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "lfontanills@gmail.com"
linkedin: "lfontanills"
github: "lfontanills"
home: "lfontanills.github.io"
# !!! You need to provide a logo image here !!! Or just delete the field for no logo
# logo: "logo_gallery.png"
output:
  epuRate::epurate:
    toc: TRUE
    number_sections: FALSE
    code_folding: "hide"
---


<br><br>

<!-- > Let's start with a short introduction. Explain briefly what this document is going to talk about. May be add a [useful link](https://github.com/holtzy/epuRate) relative to this project. -->

> In this document I will import, clean, and analyze 12 months of bike-share user data. My task is to analyze ride data to determine how annual members and casual riders use the bike-share differently. This analysis will guide my recommendations for marketing strategy to convert casual riders into annual members. I completed this case study as part of my Google Data Analytics Certification. Check out the [full project repository](https://github.com/lfontanills/bikeshare-case-study) to see how I processed the 12 data sets and created a data frame ready for analysis.

# Get the data
***
<!-- Load libraries -->

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse) # data processing and analysis
library(lubridate) # wrangle dates and times
library(kableExtra) # generate data tables
library(rmarkdown) # You need this library to run this template.
library(epuRate)  # Install with devtools: install_github("holtzy/epuRate", force=TRUE)

```

> The case study is about a fictional bike-share company called Cyclistic, while the data is taken from Chicago's Divvy bike-share, so in this report I will describe some considerations and assumptions I made to reconcile the case and the data set.

<!-- Comparison table: Cyclistic vs. Divvy and things to consider -->
```{r comparison_table}

comparison_table <- tibble(
  Item = c("Member types", "Bike types", "Docking", "Pricing"),
  Cyclistic = c("Member, Casual", "Classic, E-bike, Accessible bikes", "Must be docked", "Unspecified"),
  Divvy = c("Member, Casual", "Classic, E-Bike, Docked bike", "Docked, undocked with fees", "Flat rates with fees")
)

comparison_table %>% 
  kbl() %>% 
  kable_material(c("striped"))
```

> * Bike types: Since Cyclistic's accessible bikes account for 8% of riders, and Divvy does not have accessible bikes, I consider only classic and e-bikes in my analysis.
* Docking: Divvy docks use a different geolocation system than the bikes themselves-- docks have geolocation with 0.000001° sensitivity, while undocked bikes have geolocation with 0.01° sensitivity. Many Divvy docks are within 0.01° of other docks, especially in the city center where most rides take place, so undocked bikes can't reliably be matched with nearby docks. As Cyclistic bikes must all be docked, I am considering only bikes that start and end at docks in my analysis.

<!-- Then I explain what is my input data and I load it. Here is how it looks like: --> 

> My data for this project is located in 12 csvs, one for each of the last 12 months. I reviewed each month independently in Excel, which helped me uncover some issues to fix. These issues were:
* A test ride outside the Chicago area, and another few test rides with no geolocation
* Several rides starting before and ending after the DST time changes in March and November
* Many rides shorter than 1 minute, and longer than 24 hours
* Rides with no end location
* A nonspecific "docked" bike type assoiated with some rides by casual members
I decided to filter out these rides to better map my data to the Cyclistic case. 

```{r read df}
rides_all <- read.csv("rides_all_clean.csv")
head(rides_all)
```

This file has `r nrow(rides_all)` lines and `r ncol(rides_all)` columns. It is ready to be analyzed.

# Analysis
***
<!-- Here are 3 version of a [circular barplot](https://www.r-graph-gallery.com/circular-barplot/) that you can find in the [R graph gallery](https://www.r-graph-gallery.com/circular-barplot/). Note that the figure is centered. You can change the size of the plot with **fig.width** and **fig.height**   -->

## Observation 1: Casual riders ride more often during summer
Let's start with a basic version of a circular representation of the data
```{r, warning=FALSE, fig.align='center' }

```

## Observation 2: Casual riders use the same few stations
```{r, warning=FALSE, fig.align='center' }

```

## Observation 3: Casual riders use bikes more often on off-peak times, and for longer
```{r, warning=FALSE, fig.align='center' }

```

# Conclusion
***


## Even more stuff
***

```{r}

```


## References
***


<!-- Read more about this template [here](https://github.com/holtzy/epuRate). -->

<!-- Learn how to custom your R Markdown document [here](https://holtzy.github.io/Pimp-my-rmd/). -->
