---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Laura Fontanills"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "lfontanills@gmail.com"
linkedin: "lfontanills"
github: "lfontanills"
home: "lfontanills.github.io"
# !!! You need to provide a logo image here !!! Or just delete the field for no logo
# logo: "logo_gallery.png"
output:
  epuRate::epurate:
    toc: TRUE
    number_sections: FALSE
    code_folding: "hide"
---


<br><br>

<!-- > Let's start with a short introduction. Explain briefly what this document is going to talk about. May be add a [useful link](https://github.com/holtzy/epuRate) relative to this project. -->

> In this document I will import, clean, and analyze 12 months of bike-share user data. My task is to analyze ride data to determine how annual members and casual riders use the bike-share differently. This analysis will guide my recommendations for marketing strategy to convert casual riders into annual members. I completed this case study as part of my Google Data Analytics Certification. Check out the [full project repository](https://github.com/lfontanills/bikeshare-case-study) to see how I processed the 12 data sets and created a data frame ready for analysis.

# Get the data
***
<!-- Load libraries -->

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse) # data processing and analysis
library(lubridate) # wrangle dates and times
library(kableExtra) # generate data tables
library(rmarkdown) # You need this library to run this template.
library(epuRate)  # Install with devtools: install_github("holtzy/epuRate", force=TRUE)

```

> The case study is about a fictional bike-share company called Cyclistic, while the data is taken from Chicago's Divvy bike-share, so in this report I will describe some considerations and assumptions I made to reconcile the case and the data set.

<!-- Comparison table: Cyclistic vs. Divvy and things to consider -->
```{r comparison_table}

comparison_table <- tibble(
  Item = c("Member types", "Bike types", "Docking", "Pricing"),
  Cyclistic = c("Member, Casual", "Classic, E-bike, Accessible bikes", "Must be docked", "Unspecified"),
  Divvy = c("Member, Casual", "Classic, E-Bike, Docked bike", "Docked, undocked with fees", "Flat rates with fees")
)

comparison_table %>% 
  kbl() %>% 
  kable_material(c("striped"))
```

> * Bike types: Since Cyclistic's accessible bikes account for 8% of riders, and Divvy does not have accessible bikes, I consider only classic and e-bikes in my analysis.
* Docking: Divvy docks use a different geolocation system than the bikes themselves-- docks have geolocation with 0.000001° sensitivity, while undocked bikes have geolocation with 0.01° sensitivity. Many Divvy docks are within 0.01° of other docks, especially in the city center where most rides take place, so undocked bikes can't reliably be matched with nearby docks. As Cyclistic bikes must all be docked, I am considering only bikes that start and end at docks in my analysis.

<!-- Then I explain what is my input data and I load it. Here is how it looks like: --> 

> My data for this project is located in 12 csvs, one for each of the last 12 months. I reviewed each month independently in Excel, which helped me uncover some issues to fix. These issues were:
* A test ride outside the Chicago area, and another few test rides with no geolocation
* Several rides starting before and ending after the DST time changes in March and November
* Many rides shorter than 1 minute, and longer than 24 hours
* Rides with no end location
* A nonspecific "docked" bike type assoiated with some rides by casual members
I decided to filter out these rides to better map my data to the Cyclistic case. 

```{r}
# Show it:
rides_2022 <- read.csv("rides_2022.csv")
head(rides_2022)
```

> 

This file has `r nrow(rides_2022)` lines and `r ncol(rides_2022)` columns. It is ready to be analysed.

# Analysis
***
<!-- Here are 3 version of a [circular barplot](https://www.r-graph-gallery.com/circular-barplot/) that you can find in the [R graph gallery](https://www.r-graph-gallery.com/circular-barplot/). Note that the figure is centered. You can change the size of the plot with **fig.width** and **fig.height**   -->

## basic
Let's start with a basic version of a circular representation of the data
```{r, warning=FALSE, fig.align='center' }
# Create dataset
data=data.frame(
  individual=paste( "Mister ", seq(1,60), sep=""),
  value=sample( seq(10,100), 60, replace=T)
)

# Set a number of 'empty bar'
empty_bar=10

# Add lines to the initial dataset
to_add = matrix(NA, empty_bar, ncol(data))
colnames(to_add) = colnames(data)
data=rbind(data, to_add)
data$id=seq(1, nrow(data))

# Get the name and the y position of each label
label_data=data
number_of_bar=nrow(label_data)
angle= 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust<-ifelse( angle < -90, 1, 0)
label_data$angle<-ifelse(angle < -90, angle+180, angle)

# Make the plot
p = ggplot(data, aes(x=as.factor(id), y=value)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  geom_bar(stat="identity", fill="#69b3a2") +
  ylim(-100,120) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm")
  ) +
  coord_polar(start = 0) +
  geom_text(data=label_data, aes(x=id, y=value+10, label=individual, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE )

p
```


## grouped
```{r, warning=FALSE, fig.align='center' }
# Create dataset
data=data.frame(
  individual=paste( "Mister ", seq(1,60), sep=""),
  group=c( rep('A', 10), rep('B', 30), rep('C', 14), rep('D', 6)) ,
  value=sample( seq(10,100), 60, replace=T)
)

# Set a number of 'empty bar' to add at the end of each group
empty_bar=4
to_add = data.frame( matrix(NA, empty_bar*nlevels(data$group), ncol(data)) )
colnames(to_add) = colnames(data)
to_add$group=rep(levels(data$group), each=empty_bar)
data=rbind(data, to_add)
data=data %>% arrange(group)
data$id=seq(1, nrow(data))

# Get the name and the y position of each label
label_data=data
number_of_bar=nrow(label_data)
angle= 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust<-ifelse( angle < -90, 1, 0)
label_data$angle<-ifelse(angle < -90, angle+180, angle)

# Make the plot
p = ggplot(data, aes(x=as.factor(id), y=value, fill=group)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  geom_bar(stat="identity", alpha=0.5) +
  ylim(-100,120) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm")
  ) +
  coord_polar() +
  geom_text(data=label_data, aes(x=id, y=value+10, label=individual, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE )

p
```

## complex
```{r, warning=FALSE, fig.align='center' }
# Create dataset
data=data.frame(
  individual=paste( "Mister ", seq(1,60), sep=""),
  group=c( rep('A', 10), rep('B', 30), rep('C', 14), rep('D', 6)) ,
  value=sample( seq(10,100), 60, replace=T)
)

# Set a number of 'empty bar' to add at the end of each group
empty_bar=3
to_add = data.frame( matrix(NA, empty_bar*nlevels(data$group), ncol(data)) )
colnames(to_add) = colnames(data)
to_add$group=rep(levels(data$group), each=empty_bar)
data=rbind(data, to_add)
data=data %>% arrange(group)
data$id=seq(1, nrow(data))

# Get the name and the y position of each label
label_data=data
number_of_bar=nrow(label_data)
angle= 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust<-ifelse( angle < -90, 1, 0)
label_data$angle<-ifelse(angle < -90, angle+180, angle)

# prepare a data frame for base lines
base_data=data %>%
  group_by(group) %>%
  summarize(start=min(id), end=max(id) - empty_bar) %>%
  rowwise() %>%
  mutate(title=mean(c(start, end)))

# prepare a data frame for grid (scales)
grid_data = base_data
grid_data$end = grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
grid_data$start = grid_data$start - 1
grid_data=grid_data[-1,]

# Make the plot
p = ggplot(data, aes(x=as.factor(id), y=value, fill=group)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar

  geom_bar(aes(x=as.factor(id), y=value, fill=group), stat="identity", alpha=0.5) +

  # Add a val=100/75/50/25 lines. I do it at the beginning to make sur barplots are OVER it.
  geom_segment(data=grid_data, aes(x = end, y = 80, xend = start, yend = 80), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 60, xend = start, yend = 60), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 40, xend = start, yend = 40), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 20, xend = start, yend = 20), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +

  # Add text showing the value of each 100/75/50/25 lines
  annotate("text", x = rep(max(data$id),4), y = c(20, 40, 60, 80), label = c("20", "40", "60", "80") , color="grey", size=3 , angle=0, fontface="bold", hjust=1) +

  geom_bar(aes(x=as.factor(id), y=value, fill=group), stat="identity", alpha=0.5) +
  ylim(-100,120) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm")
  ) +
  coord_polar() +
  geom_text(data=label_data, aes(x=id, y=value+10, label=individual, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE ) +

  # Add base line information
  geom_segment(data=base_data, aes(x = start, y = -5, xend = end, yend = -5), colour = "black", alpha=0.8, size=0.6 , inherit.aes = FALSE )  +
  geom_text(data=base_data, aes(x = title, y = -18, label=group), hjust=c(1,1,0,0), colour = "black", alpha=0.8, size=4, fontface="bold", inherit.aes = FALSE)

p
```



# Conclusion
***
Read more about this template [here](https://github.com/holtzy/epuRate).

Learn how to custom your R Markdown document [here](https://holtzy.github.io/Pimp-my-rmd/).
