---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Laura Fontanills"
date: "`r format(Sys.time(), '%d %B %Y')`"
# !!! You need to provide a logo image here !!! Or just delete the field for no logo
# logo: "logo_gallery.png"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_collapsed: TRUE
    number_sections: FALSE
    code_folding: "hide"
---

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse) # data processing and analysis
library(lubridate) # wrangle dates and times
library(kableExtra) # generate data tables
options(dplyr.summarise.inform = FALSE)
```


# Read this first
***
If you'd like to skip all the data cleaning steps, click on "Get the data" in the sidebar. 

# About
***
I completed this case study as the capstone project for the Google Data Analytics Certification course. I used base R and tidyverse to complete my analysis and create visualizations. 

The case study is about Chicago's Divvy bike-share, and my objective evaluate how members and non-members ("casual") use the bike-share differently. I analyzed the duration of user trips, the locations of these trips, and the types of bikes used.

The data is taken from a real bike-share -- Chicago-based Divvy -- but in the original case study my objective was set for a fictional company. Because there were some differences between Divvy and the fictional Cyclistic bike-share, I decided to conduct this analysis as if I were conducting this analysis for Divvy.

# Data pre-processing
***

[Click here](https://divvy-tripdata.s3.amazonaws.com/index.html) to view historical Divvy data. Each month's data is contained in a single .csv file. I looked at data for the previous 12 months (January 2022 to December 2022).

Before working in R, I reviewed each month's data independently in Excel, to get a better sense of the data. I discovered the following inconsistencies that I would need to address.
* Test rides outside the Chicago area or with no geolocation
* Rides starting before and ending after the DST time changes in March and November
* Many rides shorter than 1 minute, which could be due to bike problems
* Many rides longer than 24 hours, which could be due to docking problems
* Rides with no end location
* A nonspecific "docked" bike type associated with a small percentage of rides by casual users

I combined the 12 months of Divvy data into a single data frame.
```{r read combined df}
rides_2022 <- read.csv("rides_2022.csv")
head(rides_2022)
```

This file has `r nrow(rides_2022)` lines and `r ncol(rides_2022)` columns. 

I confirm that all rows are unique by checking the number of ride IDs. TRUE indicates that all rows are unique.
```{r check all rides unique}
n_distinct(rides_2022$ride_id) == nrow(rides_2022)
```

I also confirm that all rides were taken by either members or casual users.
```{r check all user types}
rides_2022 %>% 
  group_by(member_casual) %>% 
  summarize(num_rides = n()) %>% 
  mutate(pct_rides = 100*(num_rides/sum(num_rides))) # calculate percentage
```

### Ride times and durations

I transform the started_at and ended_at fields from character vectors to datatime.

```{r lubridate transform started_at ended_at}
rides_2022$started_at <- ymd_hms(rides_2022$started_at)
rides_2022$ended_at <- ymd_hms(rides_2022$ended_at)
```

I create a new column for ride duration in seconds. Some rides have negative ride duration, so I'll filter these out.

```{r create ride length column}
rides_2022$ride_length <- as.numeric(rides_2022$ended_at - rides_2022$started_at)


rides_2022 %>% 
  filter(ride_length < 0) %>% 
  select(started_at, ended_at, ride_length) %>% 
  head()
```

I'll also need to filter out rides that started during daylight savings time (before November 6th at 2 AM) and ended after daylight savings time (after November 6th at 2 AM). This is because on November 6th the hour from 2-3 AM happens twice. This means that some of the ride durations will be incorrect, as they should be an hour longer.

```{r find rides on nov 6 - dst}
rides_2022 %>% 
  filter(started_at < '2022-11-06 02:00:00' & ended_at > '2022-11-06 02:00:00') %>% 
  select(started_at, ended_at, ride_length) %>% 
  head()
```


`r nrow(rides_2022[rides_2022$ride_length == 60, ])` rides are very short (less than 60 seconds), and of these `r nrow(rides_2022[rides_2022$ride_length == 0, ])` are 0 seconds long.


`r nrow(rides_2022[rides_2022$ride_length > (60*60*24), ])` rides are very long (greater than 24 hours). With a day pass, Divvy lets you ride up to 3 hours on a single trip before incurring a fee, and the pass is good for 24 hours. It's impossible to know whether trips longer than 3 hours were real trips or due to docking issues. `r nrow(rides_2022[rides_2022$ride_length > (60*60*3), ])` rides are longer than 3 hours. For my analysis I will consider typical use to be between 60 seconds and 3 hours.


### Ride locations and bike types

There are three bike types in rides_2022 - classic_bike, electric_bike, and docked_bike. Divvy docks use a different geolocation system than the bikes themselves-- docks have geolocation with 0.00001° sensitivity, while undocked bikes have geolocation with 0.01° sensitivity. Many Divvy docks are within 0.01° of other docks, especially in the city center where most rides take place, so undocked bikes can't reliably be matched with nearby docks. Divvy allows their users to take out and drop off undocked bikes for a fee.

```{r num rides vs bike type}
# How many rides were taken by members, casual users on each type
rides_2022 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarize(num_rides = n()) %>% 
  mutate(pct_rides = 100*(num_rides/sum(num_rides))) # calculate percentage
```

`r nrow(rides_2022[rides_2022$rideable_type == "docked_bike", ])` rides were taken on the nonspecific docked_bike type. Only casual riders used the docked_bike type, which accounted for 7.6% of all casual rides. These bikes were undocked from 678 unique stations during all months of the year. 

```{r inspect docked bike rideable type}

# When were docked bikes taken out, returned
rides_2022 %>% 
  filter(rideable_type == "docked_bike") %>% 
  group_by(rideable_type) %>% 
  summarize(min_start = min(started_at), 
            max_start = max(started_at),
            min_end = min(ended_at),
            max_end = max(ended_at))

# Where were docked bikes taken out
rides_2022 %>% 
  filter(rideable_type == "docked_bike") %>% 
  group_by(start_station_name) %>% 
  summarize(num_rides=n()) %>% 
  # arrange(num_rides)
  arrange(desc(num_rides))
```


All rides had starting geolocations in the start_lat and start_lng fields. `r nrow(rides_2022[rides_2022$start_station_name == "", ])` rides had no starting station information. All of these bikes were electric_bike type.
`
Not all rides had starting geolocations in the end_lat and end_lng fields. `r nrow(rides_2022[rides_2022$end_station_name == "", ]) # count rides w/o start station name` rides had no ending station information. These bikes were classic_bike and electric_bike types.

I also located 1 ride outside the city limits, which I identified as a test ride, and 8 rides to Green St & Madison Ave, which is not listed as a station (current or former) on Divvy's website. I will remove these from the data set during cleaning.
```{r ride locations}
# set city limits
city_lat <- c(41, 43)
city_lng <- c(-89, -87)

# find rides outside city area
rides_2022 %>% 
  filter((!between(start_lat,  city_lat[1], city_lat[2])) | 
           (!between(start_lng,  city_lng[1], city_lng[2]))) %>% 
  select(ride_id, start_station_name, start_lat, start_lng)

rides_2022 %>% 
  filter((!between(end_lat,  city_lat[1], city_lat[2])) | 
           (!between(end_lng,  city_lng[1], city_lng[2])) & end_lat != 0 & end_lng != 0) %>% 
  select(ride_id, end_station_name, end_lat, end_lng)

```

Let's clean up this data set!

```{r make clean data frame}

# create new data frame - clean
rides_clean <-rides_2022 %>% 
  filter(ride_length > 60 & ride_length < 60*60*3) %>% 
  filter(!is.na(start_lat)) %>% 
  filter(!is.na(end_lat)) %>% 
  filter(!(started_at < '2022-11-06 02:00:00' & ended_at > '2022-11-06 02:00:00')) %>% 
  filter(between(start_lat, city_lat[1], city_lat[2])) %>% 
  filter(between(end_lat, city_lat[1], city_lat[2])) %>% 
  select(ride_id, member_casual, rideable_type, started_at, ended_at, start_station_name, end_station_name, start_lat, start_lng, end_lat, end_lng, ride_length)

head(rides_clean)
```

# Get the data

```{r read rides_clean}
rides_clean <- read.csv("rides_clean.csv")
head(rides_clean)
```

This file has `r nrow(rides_clean)` lines and `r ncol(rides_clean)` columns. It is ready to be analyzed.

# Analysis and Visualizations

## Finding 1: Casual users take longer trips than members

Casual trips are about 8 minutes longer, on average, than member trips. Perhaps they're not as used to bikes or bike-shares, or perhaps they are using the bikes for fun.

```{r get summary statistics for duration}
tapply(rides_clean$ride_length, rides_clean$member_casual, summary)
```

```{r plot durations, warning=FALSE, fig.align='center' }

length_plot <-
rides_clean %>% 
  group_by(member_casual) %>% 
  summarize(number_of_rides = n(), mean_length = mean(ride_length)) %>% 
  arrange(member_casual) %>% 
  ggplot(aes(x=member_casual, y=mean_length, fill=member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average ride duration",
    subtitle = "For all trips between 1 minute and 24 hours long",
    x = "User type", 
    y = "Average ride duration (seconds)",
    fill = "User type") +
  scale_y_continuous(labels = scales::comma) +
  theme_bw() + 
  scale_fill_brewer(palette = "Greens", direction = -1)

length_plot
```

## Finding 2: Casual riders take more weekend trips

Member rides peak during the work week, which could indicate that many members use bikes for commuting. Casual rides peak on Saturday and Sunday, suggesting that casual users use bikes for leisure or tourism.

```{r plot rides by weekday, fig.align='center', warning=FALSE}
weekday_plot <-
  rides_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE, abbr = FALSE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(count=n()) %>% 
  ggplot(aes(x=weekday, y=count, fill=member_casual))+
  geom_col(position="dodge") +
  labs(
    title = "Number of rides by weekday",
    subtitle = "For all trips between 1 minute and 24 hours long",
    x = "User type", 
    y = "Number of rides",
    fill = "User type") +
  scale_y_continuous(labels = scales::comma) +
  theme_bw() + 
  scale_fill_brewer(palette = "Blues", direction = -1)

weekday_plot
```

## Finding 3: Casual riders use a few stations relatively frequently

Casual riders show a clear preference for 5 stations in particular -- these stations make up about 8% of all casual rides (by starting point). The top 5 casual starting stations are located in parks in the city center, close to the waterfront. These are very popular locations for tourists and locals.

Members don't especially favor any stations, and no station accounts for more than 1% of all member rides (by starting point).

If you'd like to see a fancy map animation I made with Divvy rider data, check out the Visualizations section of my website.

```{r plot rides by member type, warning=FALSE, fig.align='center' }

  rides_clean %>% 
  filter(member_casual == "casual") %>% 
  group_by(member_casual, start_station_name) %>% 
  summarize(num_rides = n()) %>% 
  arrange(desc(num_rides)) %>% 
  mutate(percent_rides = 100*(num_rides/sum(num_rides))) %>% 
  head(10)

  rides_clean %>% 
  filter(member_casual == "member") %>% 
  group_by(member_casual, start_station_name) %>% 
  summarize(num_rides = n()) %>% 
  arrange(desc(num_rides)) %>% 
  mutate(percent_rides = 100*(num_rides/sum(num_rides))) %>% 
  head(10)
```
## Finding 4: Casual riders use more e-bikes

In 2022, Casual riders used more electric bikes than classic bikes. By contrast, members favored classic bikes.

```{r summary bike types}
rides_clean %>% 
  group_by(member_casual, rideable_type) %>%
  summarize(num_rides = n()) %>% 
  mutate(pct_rides = 100* num_rides/sum(num_rides))
```


```{r bike types plot,warning=FALSE, fig.align='center' }
bike_type_plot <- rides_clean %>% 
  group_by(rideable_type, member_casual) %>% 
  filter(rideable_type == "classic_bike" | rideable_type == "electric_bike") %>% 
  summarize(num_rides = n()) %>% 
  arrange(member_casual) %>% 
  ggplot(aes(x=member_casual, y=num_rides, fill=rideable_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of rides with Classic Bikes, Electric Bikes",
    subtitle = "For all trips between 1 minute and 3 hours long",
    x = "Membership type", 
    y = "Number of rides",
    fill = "Bike type", 
    labels = c("classic_bike", "electric_bike")) +
  scale_y_continuous(labels = scales::comma) +
  theme_bw() + 
  scale_fill_brewer(palette = "Blues")

bike_type_plot
```


# Conclusion

Here are my recommendations for converting casual riders to members:
* Find ways to make membership more appealing to e-bike users
* Market to non-commuter casual users who use bikes for leisure
* Increase advertising at high-traffic stations in the city center

## Some extras

Here's a sample of the statistical tests I used to determine whether casual users differed significantly from members. 

```{r ride duration stats, warning=FALSE, fig.align='center' }

# Ride durations between members and casual users
ggplot(rides_clean, aes(x = ride_length)) +
  geom_histogram() +
  facet_wrap(~ member_casual) +
  labs(
    x = "ride length", 
    y = "number of rides") +
  scale_y_continuous(labels = scales::comma)

users_length <- subset(rides_clean, member_casual = 'member', select=c('member_casual', 'ride_length'))

wilcox.test(users_length$ride_length ~ users_length$member_casual)
```


## Resources
***
[Data source](https://divvy-tripdata.s3.amazonaws.com/index.html)

[Data license](https://ride.divvybikes.com/data-license-agreement)

[City of Chicago Divvy database](https://data.cityofchicago.org/Transportation/Divvy-Bicycle-Stations-All-Map/bk89-9dk7)
